Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-06-10 draft.tex
Cleaned up comments in the blind submission. Added authors and acknowledgements.

%% indicate IEEE Member or Student Member in form indicated below
\author{Submission 194}
\authorfooter{
%% insert punctuation at end of each item
% \item
%  Otto-v.-Guericke-University Magdeburg
}

@@

%% Abstract section.
\abstract{%
%(The problem).
Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease-specific risk factors.
%
Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
%
The data are analyzed using a priori defined hypotheses to find statistically significant correlations between features (\emph{associations}).
%
Modern cohort studies incorporate medical image data.
%
%Analyzing these data requires image segmentation, extraction of key figures, and shape-based subject grouping.
%
The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics, such as distances extracted from landmarks of the segmentation model.
\\\\
%(New Solution).
We propose an Interactive Visual Analysis (\emph{IVA}) approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic features and attributes derived from the image data.
%
This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as association measures.
%incorporating coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
%Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
%(Validation).
The presented concepts are applied expert-guided to gather and evaluate hypotheses about the aging process of the lumbar spine.
%
%(Results). 
It allows a more flexible comparison between image and non-image data.
%
%(Implications).
The new framework enables hypotheses validation and generation by incorporating human pattern recognition as well as data mining methods.
%
Using all reliable information from the image segmentation linked to non-image features aims to unveil \emph{associations} by applying an iterative analysis approach.

%
%Similarity measures between data variables are used to compute interesting changes in variable associations for the current variable selection.
%
%Shape-based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
} % end of abstract

% \abstract{Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease specific risk factors.
% %
% Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
% %
% %Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations (\emph{associations}).
% Driven by a priori defined hypotheses, the data is analyzed using well established statistical methods in order to find statistically significant correlations between variables (\emph{associations}).
% %
% Modern cohort studies also incorporate medical image data.
% %
% Analyzing these data requires image segmentation, extraction of key figures, and shape-based subject grouping.
% %
% The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
% \\\\
% We propose an Interactive Visual Analysis approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
% %
% Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{associations} by applying an iterative analysis approach.
% %As extension to the epidemiological workflow it aims to unveil associations based on all reliable shape information rather than abstracting it to a few key figures.
% %Iterative analysis between ... allows for extracting associations which without the visiual analysis driven approach maybe would be overlooked. %Einsicht in Interaktionen zwischen nicht-bilddaten und bilddaten auf basis aller verfügbarer Informationen 
% %Iterative Analyse von Organstrukturen im Bezug auf Bilddaten 
% %Iterative analysis of image-based epidemiological datasets aims to provide insight into associations between organ shape and % in order to provide insights into \emph{associations} by incorporating additional information
% %It extends the epidemiological workflow with an iterative Analysis process and 
% It allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
% %
% Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
% %
% Similarity measures between data variables are used to compute interesting changes in variable associations for the current variable selection.
% %
% Shape-based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
% } % end of abstract

@@

%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims at characterizing health and disease by determining risk factors.
%
Clinical problems, such as the selection of diagnostic tools and efficient treatment, are tackled using results of epidemiological research.
%
Also, the introduction of preventive measures in medicine and beyond is based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
%
Observations made by clinicians in the daily routine are translated into hypotheses for epidemiological research.
%
These are used to determine environmental and lifestyle factors as well as medical examination results that may influence a disease.
%
Potentially useful data variables (\emph{features}) are gathered using structured interviews and clinical examinations.
%
Methods like regression analysis are employed to check the attribute list for statistical soundness.
%
\\\\
Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The cohort consists of several thousands of persons randomly selected to avoid any bias.
%
Due to this random selection, the majority of the cohort does not suffer from a specific health problem.
%
A large size of the cohort is therefore essential to investigate differences between healthy and diseased people.
%
% The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
% %
% When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
% %
% Therefore results need to be age-adjusted to be comparable.
% %
% Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
%
Cohort studies often include medical image data.
%
%These data needs to be segmented to analyze its features for correlations with non-image data.
Segmentation of these data would allow for anatomical structure analysis and enable correlation with other features.
%The segmentation of these data would allow to analyze shapes of anatomical structures and to correlate them with other features.
%These data needs to be concurrently analyzed with the non-image data.
%
%This requires their segmentation and the derivation of features such as kidney volume, curvature of the spine or tissue texture of the liver.
%
%Manual segmentation is not feasible since it leads to a high variability which is at odds with the high reproducibility required in epidemiology research.
%
Semi-automatic techniques are more promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
%
Analyzing spatial data with respect to other epidemiological factors requires techniques that reach beyond standard statistical methods.

Compiling a list of features for tests of statistical resilience based on experience-driven hypotheses leaves out other features in the data which potentially interact with a disease.
%
This also applies to the chosen landmarks used to quantify medical image data information.
%
The standard workflow lacks methods, which identify correlations that the epidemiologists did not consider.
%
%Also, statistics tools are limited 
Also, only a small subset of variables can be concurrently analyzed, depending on the statistical measure.
%The hypothesis driven analysis approach of these data based on expectations of epidemiologists leaves out features which potentially interact with a disease outside.
%The hypothesis driven analysis approach of these data leaves out features which potentially interact with a disease because they they are it is outside of the erwartungshaltung of the epidemiologists.
%leaves out features which potentially interact with a disease because they are usually outside of the practical experience of the epidemiologists.
%
%Analyzing these data for correlating features requires preprocessing steps as segmentation of the structures.
%
%Manual segmentation is not feasible since it leads to a high variability which is att odds with the high reproducibility required in epidemiology research. 
%
%Semi-automatic techniques are more promising but challenging since MRI and ultrasound are subject to inhomogeneity and noise
%Due to the lack of sophisticated segmentation algorithms this is often achieved by manually specifying landmarks by radiologists which is costly and prone to inter- and intra observer variability.
%
% Since it is unethical to expose people to radiation, non-harming imaging such as Magnetic Resonance Imaging (MRI) or Ultrasound Imaging is used.
% %
% To quantify these data it is necessary to label each voxel regarding structure affiliation (\emph{segmentation}).
% %
% Manual segmentation carried out by radiological experts is possible but very costly and prone to inter- and intra observer variability.
% %
% Segmentation algorithms allow for (semi)-automated analysis of the data but require sophisticated methods due to high inter-subject variability caused by the subject diversity.
% %
% Analyzing spatial data with respect to other epidemiological factors requires techniques which reach beyond standard statistical methods.
\\\\
%We propose an Interactive Visual Analysis (\emph{IVA}) approach \cite{Thomas2005} to provide a way to analyze image and non-image data.
We propose an Interactive Visual Analysis (\emph{IVA}) approach \cite{Thomas2005} for the combined analysis of image and non-image data.
%
Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space incorporating many different variables.
%
Intended as an extension to the well-established epidemiological tools it provides a way to rapidly validate hypotheses and to trigger \emph{hypothesis generation} using data mining methods, such as clustering.
%
\emph{Hypothesis generation} gains importance since the number of epidemiological features increases and the focus shifts towards more complex relations involving more than two features.
%
The easy exchange of developed methods driven by modern web technologies intends to trigger a fast feedback loop between us and the epidemiologists.
% http://wiki.infowiss.net/Gulf_of_execution ?
We applied our approach to a data set compiled to analyze diseases related to the lumbar spine and aim to determine features, which indicate pathological changes.
%
This data set comprises 127 features and 2 sequences of MRI data from 2,333 patients.
\\\\
Our contributions are:
\begin{itemize}
%	\item providing an overview over the workflow for analyzing cohort study data to gain insight into the large diverse subject spaces.
	%\item an Interactive Visual Analysis workflow for analyzing image-based epidemiological data both hypothesis-driven and for hypothesis generation--based on a characterization of the standard epidemiological workflow,
	\item an Interactive Visual Analysis workflow for cohort study data to allow both, hypothesis-driven analysis and hypothesis generation--based on a characterization of the standard epidemiological workflow,
	%\item an Interactive Visual Analysis workflow for analyzing image-based epidemiological data including both hypothesis-driven and for hypothesis-generation based on a characterization of the standard epidemiological workflow
	\item visualization techniques, which incorporate both information visualization and 3D rendering of organ shapes as well as combining them with epidemiological graphics and key figures,
	\item highlighting interesting subject groups and feature associations using shape-based clustering and statistical contingency measures,
	%Applying Interactive Visual Analysis toolset to image-based analysis in epidemiology with focus hypothesis generation
	%\item Applying Interactive Visual Analysis toolset to image-based analysis in epidemiology with focus hypothesis generation
	% \item Describing a Interactive Visual Analysis system aiming to support hypothesis generation .. of image-based cohort study analysis
% 		\item Characterize the problems of epidemiological context which  %Applying the Interactive Visual Analysis technique set to the epidemiological problem domain by characterizing requirements of this context.
% 	\item Applying the Interactive Visual Analysis technique set to the epidemiological problem domain by characterizing requirements of this context.
% 	\item Provide an overview over the workflow for analyzing cohort study data to gain insight into the large diverse subject spaces.
% 	\item Provide visualization techniques which combine both information visualization and 3D rendering of organ shapes as well as combining them with well known epidemiological graphics and key figures.
	\item an implementation of the presented methods in a web framework based on \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}.
\end{itemize}

%Statistical correlations derived from analyzing the data itself may be misleading since

%Include Implementation details - web based; how are image data included; what technologies are used?

@@

%In this section we give insight into the epidemiological workflow when analyzing cohort study data and derive several challenges. %study data to identify the problems we address in this paper.
In this section we describe the epidemiological workflow and associated requirements.
% give insight into the epidemiological workflow when analyzing cohort study data and derive several challenges
%
%Epidemiological Outcome - 5Ds:
% Death
% Disease
% Discomfort
% Disability
% Dissatisfaction
%
\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/workflow_comparison}
 \caption{\emph{IVA} tools complement parts of the epidemiological workflow instead of replacing them. The appropriate combination of statistical and interactive-driven analysis shows promising potential to unveil the information in the data. (a) shows the standard epidemiological workflow, (b) the \emph{IVA} supported one. The iterative red highlighted part is called the \emph{IVA Loop} and is described in more detail in Figure~\ref{fig:InteractionLoop}.}
  \label{fig:WorkflowComparison}
\end{figure}
%Epidemiology diversity is reflected in the different experts working at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics, and statisticians.
%Many different experts work at epidemiological studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
%
Epidemiologists follow a workflow mainly driven by statistic tools to validate hypotheses about disease-specific risk factors.
%
Following Thew and colleagues, the workflow can be characterized as follows \cite{Thew2009}:
%
\begin{enumerate}
%	\item Hypotheses are based most commonly on observations made by clinicians in their daily routine.
	\item A hypothesis is derived from observations made by clinicians in their daily routine.
%
	\item A set of features depicting conditions affected by the hypothesis is compiled accordingly.
	%\item Confounding features are adjusted so that they do not influence the effect size of an attribute.
	\item Confounding features are identified and taken into account (for example using stratification). % adjusted so that they do not influence the effect size of a attribute.
%
	\item Statistical methods, such as regression analysis, assess the association of selected features with the investigated disease.%are applied to measure the effect size of attributes to the outcome of interest.
	%\item Statistical methods such as regression analysis are applied to measure the effect size of attributes to the outcome of interest.
\end{enumerate}
The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a) and serves as orientation for our approach although it does not consider any kind of image data.
%
We focus, on the potential of image data and attempt to support \emph{hypothesis generation}.
%The extension we suggest, is motivated by the potential of image data and the attempt to support \emph{hypothesis generation}.
%In the following we discuss the problems and challenges regarding the workflow.

Reproducibility of results is an epidemiological key requirement.
%
It is difficult to achieve, since many physicians are involved when thousands of test persons are examined and interviewed.
%
Thus, both intra- and inter-observer variability needs to be low for all aspects of a cohort study examination.
%
Longitudinal studies require the acquired attributes to be comparable for evaluation.
%
If the data acquisition process changes, an information bias is introduced to the data, hampering inference in and between acquisition cycles.
%
%This underlines the high quality standards to methods processing the data, whether to extract additional parameters or gain insight.

In longitudinal cohort studies, grouping subjects using epidemiological features is essential in order to allow per-group risk determination.
%
%Grouping is carried out hypothesis driven.
Grouping depends on the underlying hypothesis.
%
Age for example is divided into groups (e.g. in 20 year steps) when investigating its influence on a condition.
%
These groups strongly depend on the condition of interest and therefore there is no standard on how to categorize them.

To determine whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values, which indicate statistical significance.
%
Statistical correlations are prone to \emph{confounding}, meaning that the association of two features is influenced by a third feature, which needs to be isolated.
%
A famous example for a confounder is the association between shoe size and mortality, where it can be observed that people with larger shoe size have a smaller life expectation than people with small shoe size.
%
The shoe size is actually associated with gender, where women have smaller feet than men but also a longer life expectation.
% Statistical correlations are prone to \emph{confounding}, meaning that two features are dependent and therefore should be normalized with respect to each other.
%
%Statistics tools such as \texttt{SPSS}\footnote{Product of IBM; \url{www.ibm.com/software/de/analytics/spss/}} and \texttt{Stata}\footnote{Product of Stata Corp.; \url{http://www.stata.com/}} play a major role for analyzing epidemiological data.

@@

\subsection{Epidemiological Data} \label{EpidemiologicalData}
Epidemiological data are highly heterogenous and incomplete.
%
Information about medical history and examinations, genetic conditions, geographical data, questionnaire results and image data yield a complex data space for each subject.
%
%Often data are derived from acquired variables to either group or threshold values or to get information derived from reviewed data such as breast density data for women.
%
For ethical, legal or medical reasons some features cannot be gathered for each subject.
%
The most obvious example are women-specific questions about menstrual status or number of born children.
%
Follow-up examinations or questions about conditions like medications taken after a diagnosed disease also yield features only available for a small amount of subjects.
%

Indicators for medical conditions as well as questions about a subject's lifestyle are also often \emph{dichotomous}--they have two manifestations (\emph{Yes} or \emph{No}).
%
%This allows for the calculation of \emph{odds ratios} which describe the relation of two \emph{dichotomous} variables, allowing for direct comparison of their influences.
%
Dichotomous data can also be derived by aggregating features to yield only two manifestations (e.g. subjects younger or older than 50 years).
%
Medical examinations mostly comprise categorical (e.g. levels of back pain) and continuous values (e.g. age or body size).

\paragraph{Image acquisition.} \label{ImageAcquisition} Imaging techniques involving ionizing radiation for the subject are not suitable for ethical reasons.
%
%MRI data is more expensive to obtain than CT data but does not affect the subjects health and is therefore the main method for collecting cohort study imaging data.
Therefore, MRI is the main method for collecting cohort study imaging data.
%
The image quality is a tradeoff between accuracy and affordability \cite{Preim2014}.
%
This often yields image resolutions inferior to those of clinical day-to-day practice, which makes their analysis more challenging.
%The equipment used to gather medical image data is kept, if possible, on the initial software and hardware version to ensure comparability in and between acquisition cycles.
The equipment used to gather medical image data is kept, if possible, on the initial software and hardware version during a longitudinal study to ensure comparability in and between acquisition cycles.

\paragraph{Image analysis.} \label{ImageAnalysis} Decisions have to be made on how image data are \emph{compared} and \emph{quantified}.
%
Segmentation masks representing the voxels of an anatomical structure would be ideal, since many different key figures, e.g., volume, largest diameter or aspect ratio, can be derived from them.
%
Since reliable and efficient segmentation techniques for these data are not available in general, epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and the number of subjects.
%
Information derived by landmarks, such as top and bottom point of a vertebra, are by far not as expressive and versatile as segmentation masks describing its whole shape.
%Information derived by landmarks, such as top and bottom point, are by far not as expressive and versatile as segmentation masks describing the whole vertebra.
%Information derived by landmarks such as top and bottom point of a vertebra are also not nearly as expressive and versatile as segmentation masks describing the whole vertebra.
%
They are also prone to a high inter-observer variability and are hard to reproduce.
%
%This gains even more momentum when analyzing multiple time steps.
This gains even more importance when analyzing multiple time steps.
%
Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey value distribution in an area (used for determining the type of tissue).

@@

\subsection{The Study of Health in Pomerania (SHIP)}
After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives have evolved.
%
They slightly differ in clinical focus, acquired data and epidemiological research questions.
%
Starting in 1997 with a cohort consisting of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
Data is collected without focus on a group of diseases.
%
This allows the data set to be queried regarding many different diseases and conditions.
%
Subjects were examined in a 5-year time span, continuously adding new parameters including MRI scans in the last iteration \cite{Hegenscheid2009}.
%
The MRI protocol features a rich number of sequences.
%
%In addition to other MRI data, breast MRI scans were acquired for women.
%
%A second cohort \texttt{SHIP-Trend} was established in 2008 to acquire data about a younger population.
A second cohort \texttt{SHIP-Trend} was established in 2008.
%
The protocols for examining the subjects between \texttt{SHIP} and \texttt{SHIP-Trend} remained the same, making them comparable.
%
The overall examination time for each person attending the study is two days.

\section{Prior and Related Work}
%Einfuehren von Helwigs Terminologie?
Designing a visualization, which conveys all data aspects equally, is challenging.
%
Given the number of features of epidemiological data sets and their different manifestations, the strength of different visualization techniques needs to be combined \cite{Buja91, Konyha2009}.
%
The principal component analysis (PCA) and similar techniques are able to reduce the dimensions by extracting most expressive components, but make the influence of each variable hard to convey.
\\\\
Their focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes the work of Turkay and colleagues closest to ours albeit our emphasis on processing models derived from medical image data segmentation and variables with categorical manifestations \cite{Turkay2013}.
%
Their methods aim to amplify a hypothesis generation process for analyzing data of a Norwegian aging study.
%
%Statistical measures of continuous variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures.
Statistical measures of continuous variables, such as mean, standard deviation, skewness, or inter-quartile range, are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures, such as voxel number of a segmented structure.
%
%These transform dimensions into data points and make them comparable with respect to the derived descriptive measures, making them comparable.
%
%This not only allows for comparing all continuous variables in a single plot but make their distribution change comprehensible.
%
%This requires a good descriptive measure which captures the kind of change the user is interested in or which reflects unexpected data behavior.
%
%The technique was applied to features obtained by segmenting the brain into 45 parts and measure the voxel number, volume and properties of the intensity values.
%
The method is strongly dependent on the descriptive measures of the epidemiological factors.

@@

% TODO Bernhard fragen, was ihn an der Reihenfolge stört
It is well known that overweight is a risk factor for back pain.
%
While the BMI is a key figure for assessing height and weight of a subject, it does not tell anything about how the weight is distributed in the body.
%
Our epidemiological collaborators were interested in the correlation with waist circumference presented by this group.
%Our epidemiological collaborators were interested that this group presented a correlation with waist circumference.
%
Our finding follows the recent trends indicating that BMI is not a good measure for assessing the body shape, since healthy weight is dependent on many other measures \cite{Ahima2013}.
%
It indicates that the waist circumference rather than the BMI interacts with unusually shaped spines for subjects with lumbar back pain.
%
The influence of the parameter is now in the focus of further analysis.

>>>>

%% indicate IEEE Member or Student Member in form indicated below
\author{P. Klemm$^{1}$, K. Lawonn$^{1}$, M. Rak$^{1}$, B. Preim$^{1}$, K. Toennies$^{1}$, K. Hegenscheid$^{2}$, H. Völzke$^{2}$, S. Oeltze$^{1}$}
\authorfooter{
%% insert punctuation at end of each item
\item
 $^1$Otto-von-Guericke University Magdeburg, Germany.
\item
 $^2$Ernst-Moritz-Arndt-University Greifswald, Germany.
}

@@

%% Abstract section.
\abstract{%
%(The problem).
Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease-specific risk factors.
%
Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors and more recently, medical image data.
%
%The data are analyzed using a priori defined hypotheses to find statistically significant correlations between features (\emph{associations}).
%
%Modern cohort studies incorporate medical image data.
%
%
%The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics, such as distances extracted from landmarks of the segmentation model.
%\\\\
%(New Solution).
We propose an Interactive Visual Analysis (\emph{IVA}) approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic features and attributes derived from the image data.
%
This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as association measures.
%(Validation).
The presented concepts are applied expert-guided to gather and evaluate hypotheses about the aging process of the lumbar spine.
%
%(Results). 
It allows a more flexible comparison between image and non-image data.
%
%(Implications).
The new framework enables hypotheses validation and generation by incorporating human pattern recognition as well as data mining methods.
%
Using all reliable information from the image segmentation linked to non-image features aims to unveil \emph{associations} by applying an iterative analysis approach.
} % end of abstract

@@

%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims at characterizing health and disease by determining risk factors.
%
Clinical problems, such as the selection of diagnostic tools and efficient treatment, are tackled using results of epidemiological research.
%
Also, the introduction of preventive measures in medicine and beyond is based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
%
Observations made by clinicians in the daily routine are translated into hypotheses for epidemiological research.
%
These are used to determine environmental and lifestyle factors as well as medical examination results that may influence a disease.
%
Potentially useful data variables (\emph{features}) are gathered using structured interviews and clinical examinations.
%
Methods like regression analysis are employed to check the attribute list for statistical soundness.
%
\\\\
Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The cohort consists of several thousands of persons randomly selected to avoid any bias.
%
Due to this random selection, the majority of the cohort does not suffer from a specific health problem.
%
A large size of the cohort is therefore essential to investigate differences between healthy and diseased people.
%
% The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
% %
% When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
% %
% Therefore results need to be age-adjusted to be comparable.
% %
% Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
%
Cohort studies often include medical image data.
%
Segmentation of these data would allow for anatomical structure analysis and enable correlation with other features.
%
Semi-automatic techniques are more promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
%
Analyzing spatial data with respect to other epidemiological factors requires techniques that reach beyond standard statistical methods.

Compiling a list of features for tests of statistical resilience based on experience-driven hypotheses leaves out other features in the data which potentially interact with a disease.
%
This also applies to the chosen landmarks used to quantify medical image data information.
%
The standard workflow lacks methods, which identify correlations that the epidemiologists did not consider.
%
Also, only a small subset of variables can be concurrently analyzed, depending on the statistical measure.
\\\\
We propose an Interactive Visual Analysis (\emph{IVA}) approach \cite{Thomas2005} for the combined analysis of image and non-image data.
%
Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space incorporating many different variables.
%
Intended as an extension to the well-established epidemiological tools it provides a way to rapidly validate hypotheses and to trigger \emph{hypothesis generation} using data mining methods, such as clustering.
%
\emph{Hypothesis generation} gains importance since the number of epidemiological features increases and the focus shifts towards more complex relations involving more than two features.
%
The easy exchange of developed methods driven by modern web technologies intends to trigger a fast feedback loop between us and the epidemiologists.
% http://wiki.infowiss.net/Gulf_of_execution ?
We applied our approach to a data set compiled to analyze diseases related to the lumbar spine and aim to determine features, which indicate pathological changes.
%
This data set comprises 127 features and 2 sequences of MRI data from 2,333 patients.
\\\\
Our contributions are:
\begin{itemize}
	\item an Interactive Visual Analysis workflow for cohort study data to allow both, hypothesis-driven analysis and hypothesis generation--based on a characterization of the standard epidemiological workflow,
	\item visualization techniques, which incorporate both information visualization and 3D rendering of organ shapes as well as combining them with epidemiological graphics and key figures,
	\item highlighting interesting subject groups and feature associations using shape-based clustering and statistical contingency measures,
	\item an implementation of the presented methods in a web framework based on \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}.
\end{itemize}

@@

In this section we describe the epidemiological workflow and associated requirements.

\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/workflow_comparison}
 \caption{\emph{IVA} tools complement parts of the epidemiological workflow instead of replacing them. The appropriate combination of statistical and interactive-driven analysis shows promising potential to unveil the information in the data. (a) shows the standard epidemiological workflow, (b) the \emph{IVA} supported one. The iterative red highlighted part is called the \emph{IVA Loop} and is described in more detail in Figure~\ref{fig:InteractionLoop}.}
  \label{fig:WorkflowComparison}
\end{figure}
The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics, and statisticians.
%
Epidemiologists follow a workflow mainly driven by statistic tools to validate hypotheses about disease-specific risk factors.
%
Following Thew and colleagues, the workflow can be characterized as follows \cite{Thew2009}:
%
\begin{enumerate}
	\item A hypothesis is derived from observations made by clinicians in their daily routine.
%
	\item A set of features depicting conditions affected by the hypothesis is compiled accordingly.
%
	\item Confounding features are identified and taken into account (for example using stratification).
%
	\item Statistical methods, such as regression analysis, assess the association of selected features with the investigated disease.
%
\end{enumerate}
The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a) and serves as orientation for our approach although it does not consider any kind of image data.
%
We focus, on the potential of image data and attempt to support \emph{hypothesis generation}.

Reproducibility of results is an epidemiological key requirement.
%
It is difficult to achieve, since many physicians are involved when thousands of test persons are examined and interviewed.
%
Thus, both intra- and inter-observer variability needs to be low for all aspects of a cohort study examination.
%
Longitudinal studies require the acquired attributes to be comparable for evaluation.
%
If the data acquisition process changes, an information bias is introduced to the data, hampering inference in and between acquisition cycles.
%

In longitudinal cohort studies, grouping subjects using epidemiological features is essential in order to allow per-group risk determination.
%
Grouping depends on the underlying hypothesis.
%
Age for example is divided into groups (e.g. in 20 year steps) when investigating its influence on a condition.
%
These groups strongly depend on the condition of interest and therefore there is no standard on how to categorize them.

To determine whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values, which indicate statistical significance.
%
Statistical correlations are prone to \emph{confounding}, meaning that the association of two features is influenced by a third feature, which needs to be isolated.
%
A famous example for a confounder is the association between shoe size and mortality, where it can be observed that people with larger shoe size have a smaller life expectation than people with small shoe size.
%
The shoe size is actually associated with gender, where women have smaller feet than men but also a longer life expectation.

@@

\subsection{Epidemiological Data} \label{EpidemiologicalData}
Epidemiological data are highly heterogenous and incomplete.
%
Information about medical history and examinations, genetic conditions, geographical data, questionnaire results and image data yield a complex data space for each subject.
%
For ethical, legal or medical reasons some features cannot be gathered for each subject.
%
The most obvious example are women-specific questions about menstrual status or number of born children.
%
Follow-up examinations or questions about conditions like medications taken after a diagnosed disease also yield features only available for a small amount of subjects.
%

Indicators for medical conditions as well as questions about a subject's lifestyle are also often \emph{dichotomous}--they have two manifestations (\emph{Yes} or \emph{No}).
%
Dichotomous data can also be derived by aggregating features to yield only two manifestations (e.g. subjects younger or older than 50 years).
%
Medical examinations mostly comprise categorical (e.g. levels of back pain) and continuous values (e.g. age or body size).

\paragraph{Image acquisition.} \label{ImageAcquisition} Imaging techniques involving ionizing radiation for the subject are not suitable for ethical reasons.
%
Therefore, MRI is the main method for collecting cohort study imaging data.
%
The image quality is a tradeoff between accuracy and affordability \cite{Preim2014}.
%
This often yields image resolutions inferior to those of clinical day-to-day practice, which makes their analysis more challenging.
%
The equipment used to gather medical image data is kept, if possible, on the initial software and hardware version during a longitudinal study to ensure comparability in and between acquisition cycles.

\paragraph{Image analysis.} \label{ImageAnalysis} Decisions have to be made on how image data are \emph{compared} and \emph{quantified}.
%
Segmentation masks representing the voxels of an anatomical structure would be ideal, since many different key figures, e.g., volume, largest diameter or aspect ratio, can be derived from them.
%
Since reliable and efficient segmentation techniques for these data are not available in general, epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and the number of subjects.
%
Information derived by landmarks, such as top and bottom point of a vertebra, are by far not as expressive and versatile as segmentation masks describing its whole shape.
%
They are also prone to a high inter-observer variability and are hard to reproduce.
%
This gains even more importance when analyzing multiple time steps.
%
Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey value distribution in an area (used for determining the type of tissue).

@@

\subsection{The Study of Health in Pomerania (SHIP)}
After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives have evolved.
%
They slightly differ in clinical focus, acquired data and epidemiological research questions.
%
Starting in 1997 with a cohort consisting of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
Data is collected without focus on a group of diseases.
%
This allows the data set to be queried regarding many different diseases and conditions.
%
Subjects were examined in a 5-year time span, continuously adding new parameters including MRI scans in the last iteration \cite{Hegenscheid2009}.
%
The MRI protocol features a rich number of sequences.
%
A second cohort \texttt{SHIP-Trend} was established in 2008.
%
The protocols for examining the subjects between \texttt{SHIP} and \texttt{SHIP-Trend} remained the same, making them comparable.
%
The overall examination time for each person attending the study is two days.

\section{Prior and Related Work}
Designing a visualization, which conveys all data aspects equally, is challenging.
%
Given the number of features of epidemiological data sets and their different manifestations, the strength of different visualization techniques needs to be combined \cite{Buja91, Konyha2009}.
%
The principal component analysis (PCA) and similar techniques are able to reduce the dimensions by extracting most expressive components, but make the influence of each variable hard to convey.
\\\\
Their focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes the work of Turkay and colleagues closest to ours albeit our emphasis on processing models derived from medical image data segmentation and variables with categorical manifestations \cite{Turkay2013}.
%
Their methods aim to amplify a hypothesis generation process for analyzing data of a Norwegian aging study.
%
Statistical measures of continuous variables, such as mean, standard deviation, skewness, or inter-quartile range, are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures, such as voxel number of a segmented structure.
%
%These transform dimensions into data points and make them comparable with respect to the derived descriptive measures, making them comparable.
%
%This not only allows for comparing all continuous variables in a single plot but make their distribution change comprehensible.
%
%This requires a good descriptive measure which captures the kind of change the user is interested in or which reflects unexpected data behavior.
%
%The technique was applied to features obtained by segmenting the brain into 45 parts and measure the voxel number, volume and properties of the intensity values.
%
The method is strongly dependent on the descriptive measures of the epidemiological factors.

@@

It is well known that overweight is a risk factor for back pain.
%
While the BMI is a key figure for assessing height and weight of a subject, it does not tell anything about how the weight is distributed in the body.
%
Our epidemiological collaborators were interested in the correlation with waist circumference presented by this group.
%
Our finding follows the recent trends indicating that BMI is not a good measure for assessing the body shape, since healthy weight is dependent on many other measures \cite{Ahima2013}.
%
It indicates that the waist circumference rather than the BMI interacts with unusually shaped spines for subjects with lumbar back pain.
%
The influence of the parameter is now in the focus of further analysis.
